\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{fancyvrb}
% parskip, caption omitted (not installed in local TinyTeX)
\setlength{\parskip}{0.6em}
\setlength{\parindent}{0pt}
\DefineVerbatimEnvironment{code}{Verbatim}{fontsize=\small,frame=single,framesep=3mm,xleftmargin=2mm,xrightmargin=2mm}
\DefineVerbatimEnvironment{codetiny}{Verbatim}{fontsize=\footnotesize,frame=single,framesep=3mm,xleftmargin=2mm,xrightmargin=2mm}
\title{PO-LEU Worked Example:\\
\large Trace of the Wave-11 one-customer dry-run on Amazon customer \texttt{R\_1lztobDYoYyZWKF}}
\author{PO-LEU project}
\date{April 2026}
\begin{document}
\maketitle

\begin{abstract}
\noindent
This document is a long-form walkthrough of the PO-LEU framework using
\emph{one real customer} from the Harvard-Dataverse Amazon purchase panel
as the running example. The customer, survey-response identifier
\texttt{R\_1lztobDYoYyZWKF}, contributes nine purchase events between
2018-02 and 2022-12. We take that CSV pair, push it through every one of
the twelve pipeline stages, show the tensors that emerge, the prompt the
frozen language model sees, the outcome narratives it returns, the
embeddings of those narratives, the attribute scores they induce, the
weights the weight-net assigns to this person, the loss at event $t=0$,
the four regularizer values, four steps of optimizer updates, the six
evaluation metrics, and the four interpretability reports. The run
actually happened: it is the Wave-11 \emph{dry-run} of April 2026, with
artifacts at \texttt{/tmp/w11\_dryrun\_final/}, cached outcomes in
\texttt{outcomes\_cache/outcomes.sqlite}, and cached embeddings in
\texttt{embeddings\_cache/embeddings.sqlite}. Every number in the document
is either copied from those artifacts or derived from them; fabricated
numbers are marked explicitly. The document is pedagogical: it targets an
engineer or reviewer who wants to understand what the framework does to
one row of input, not a summary for an executive.
\end{abstract}

% ============================================================
\section*{1. Setup and customer identity}
% ============================================================

The run uses the adapter \texttt{amazon} against the one-customer fixture
\texttt{configs/datasets/amazon\_1cust.yaml}. That YAML points at
\texttt{tests/fixtures/amazon\_persons\_1.csv} and
\texttt{tests/fixtures/amazon\_events\_1.csv}: a single customer plucked
from the 5{,}027-respondent Prolific survey. The survey row identifies
\texttt{R\_1lztobDYoYyZWKF} as a 25--34 year old white male in Ohio,
living alone (household size 1), earning less than \$25{,}000, with a
high-school diploma, self-reporting no diabetes, no wheelchair, and no
tobacco, cannabis, or alcohol use. The customer buys rarely: nine
purchases over a 58-month window (2018-02-11 through 2022-12-27). The
categorical span is wide: two kitchen scales early on, a refrigerator
and a no-category miscellaneous item in 2021, a quilt in late 2022, and
a tight cluster of four computer/gaming purchases all on the same day
in December 2022. The customer is therefore low-income, single-person,
not risk-averse by our lifestyle proxy, and a \emph{bursty} buyer: one
bad fit for any demographic-only baseline but a fine fit for a
narrative model that can reason about the consequences of each
purchase.

The \texttt{n\_customers=1} driver invocation is deliberately a
\emph{dry-run}: it exists to prove the code paths end to end without
spending \$200 on a 100-customer run, and to generate real cached
artifacts the Wave-11 full run can then replay. It succeeded: the
metrics on the held-out test event (event $t = 8$, a Nintendo Switch
controller) came out to top-1 accuracy $= 1.0$, top-5 $= 1.0$, MRR
$= 1.0$, NLL $= 2.184$. Every stratified breakdown and every
regularizer number is quoted further below.

% ============================================================
\section*{2. Raw data, annotated}
% ============================================================

\subsection*{2.1 Survey row (the $c_d$ / $z_d$ source)}

The entire customer metadata is a single tab in the Prolific survey CSV:

\begin{codetiny}
Survey ResponseID : R_1lztobDYoYyZWKF
Q-demos-age        : 25 - 34 years
Q-demos-hispanic   : No
Q-demos-race       : White or Caucasian
Q-demos-education  : High school diploma or GED
Q-demos-income     : Less than $25,000
Q-demos-gender     : Male
Q-sexual-orient.   : heterosexual (straight)
Q-demos-state      : Ohio
Q-amazon-use-howmany    : 1 (just me!)
Q-amazon-use-hh-size    : 1 (just me!)
Q-amazon-use-how-oft    : Less than 5 times per month
Q-substance-use-cigarettes : No
Q-substance-use-marijuana  : No
Q-substance-use-alcohol    : No
Q-personal-diabetes     : No
Q-personal-wheelchair   : No
Q-life-changes          : (blank / NaN)
Q-sell-YOUR-data        : No
Q-sell-consumer-data    : No
Q-small-biz-use         : No
Q-census-use            : I don't know
Q-research-society      : Yes
\end{codetiny}

\subsection*{2.2 The nine events}

\begin{center}\small
\begin{longtable}{@{}c l r l l p{0.27\textwidth}@{}}
\toprule
$t$ & date & price & category & ASIN & title (truncated) \\
\midrule\endhead
0 & 2018-02-11 &  9.99 & \texttt{WEIGH\_SCALE}                & B01JTDG084 & Greater Goods Gray Food Scale \\
1 & 2018-02-15 & 10.80 & \texttt{WEIGH\_SCALE}                & B012JTJJJW & TBBSC Digital Pocket Scale \\
2 & 2021-03-17 & 17.99 & \texttt{Unknown}                     & B08RDTNVYG & \emph{(title blank)} \\
3 & 2021-03-17 & 99.99 & \texttt{REFRIGERATOR}                & B07WP34LMM & Walsh WSR17BK 1.7 cu ft Compact Refrigerator \\
4 & 2022-09-03 & 49.99 & \texttt{BEDDING\_SET}                & B089NH3Q29 & Patchwork Quilt Set Full/Queen \\
5 & 2022-12-27 &  9.99 & \texttt{HEAT\_SINK}                  & B07KDDKDNN & MHQJRH M.2 2280 SSD Heatsink \\
6 & 2022-12-27 & 89.99 & \texttt{COMPUTER\_DRIVE\_OR\_STORAGE}& B07MFZY2F2 & SAMSUNG 970 EVO Plus 1TB NVMe SSD \\
7 & 2022-12-27 & 54.00 & \texttt{PHYSICAL\_VIDEO\_GAME\_SOFTWARE}& B09TMHDVLD & Pokemon Violet (Nintendo Switch) \\
8 & 2022-12-27 & 49.99 & \texttt{VIDEO\_GAME\_CONTROLLER}     & B0BDSNHLDY & HORI Switch Split Pad Compact \\
\bottomrule
\end{longtable}
\end{center}

Three time windows are visible: February 2018 (the kitchen-scale pair),
March 2021 (refrigerator plus one untitled item), and the late-2022
cluster. Event $t = 2$ carries no title and \texttt{category} is null;
the schema's \texttt{category\_null\_fill: "Unknown"} rule overwrites it
with the literal string \verb|"Unknown"|. Events $t = 5$ through $t = 8$
are on the same day and together form a full ``Switch gaming setup +
internal SSD upgrade'' basket; a survey-only baseline would have no
signal about this burst.

Given \texttt{val\_frac = test\_frac = 0.1} and $n = 9$,
\texttt{temporal\_split} allocates
$n_{\text{test}} = \max(1, \lfloor 0.1 \cdot 9\rfloor) = 1$,
$n_{\text{val}} = \max(1, \lfloor 0.1 \cdot 9\rfloor) = 1$,
$n_{\text{train}} = 9 - 1 - 1 = 7$. The seven earliest events (food
scale through Samsung SSD) become train; event $t = 7$ (Pokemon Violet)
is val; event $t = 8$ (HORI Split Pad) is test.

% ============================================================
\section*{3. Stage-by-stage trace}
% ============================================================

The driver \texttt{scripts/run\_dataset.py} executes twelve pipeline
stages, followed by model fit, evaluation, and four reports. We walk
each stage in turn, showing input shape, output shape, and a
before-vs-after snippet for this customer.

\subsection*{3.1 Stage \textsc{load} (raw CSV ingestion)}

\textbf{Purpose.} Read the two CSVs declared on the dataset YAML, no
renames.
\textbf{Input.} Paths \texttt{tests/fixtures/amazon\_events\_1.csv} and
\texttt{tests/fixtures/amazon\_persons\_1.csv}.
\textbf{Output.} Two untouched DataFrames: \texttt{events\_raw}
($9 \times 8$, raw Amazon column names with spaces) and
\texttt{persons\_raw} ($1 \times 23$).
\textbf{Invariants.} \texttt{validate\_loaded}: required source columns
from schema \texttt{events.column\_map} keys present; no silent schema
drift.

\begin{code}
events_raw.columns
# Index(['Order Date', 'Purchase Price Per Unit', 'Quantity',
#        'Shipping Address State', 'Title',
#        'ASIN/ISBN (Product Code)', 'Category',
#        'Survey ResponseID'], dtype=object)
\end{code}

\subsection*{3.2 Stage \textsc{clean} (rename, drop, coerce, sort)}

\textbf{Purpose.} Delegate to \texttt{schema\_map.translate\_events};
rename columns to canonical
\verb|{order_date, price, quantity, state, title, asin, category, customer_id}|,
drop rows null on the \texttt{dropna\_subset}, coerce \texttt{price} to
float and \texttt{asin} / \texttt{customer\_id} to string, fill null
categories to \verb|"Unknown"|, sort by
\texttt{(customer\_id, order\_date)}.
\textbf{Input / Output.} DataFrame $9 \times 8 \to 9 \times 8$ with
canonical column names (no row attrition for this customer).
\textbf{Invariants.} \texttt{validate\_cleaned}: canonical columns
present, no negative prices, no null \texttt{asin} or
\texttt{customer\_id}, no NaT in \texttt{order\_date}, RangeIndex reset.

The effect on event $t = 2$: the raw \texttt{Category} is blank; the
\texttt{category\_null\_fill} rule overwrites it with the literal
string \verb|"Unknown"|. Nothing else changes shape-wise.

\subsection*{3.3 Stage \textsc{survey\_join}}

\textbf{Purpose.} Left-join \texttt{persons\_raw} onto cleaned events
by \texttt{customer\_id} (after renaming \texttt{"Survey ResponseID"}
on the persons side); drop all-NaN columns; drop duplicate
\texttt{customer\_id} rows on the persons side
(\texttt{keep="first"}); snake-case every persons column except the
join key.
\textbf{Input / Output.} $9 \times 8$ events plus $1 \times 23$ persons
$\to$ $9 \times (8 + 22) = 9 \times 30$.
\textbf{Invariants.} \texttt{validate\_joined}: orphan rate $\leq 5\%$
(our single customer is covered: $0\%$).

\subsection*{3.4 Stage \textsc{state\_features}}

\textbf{Purpose.} Compute seven per-event columns from the customer's
prior event history: \texttt{routine}, \texttt{novelty},
\texttt{recency\_days}, \texttt{cat\_affinity}, \texttt{cat\_count\_7d},
\texttt{cat\_count\_30d}, \texttt{brand}. A first-time
\texttt{(customer\_id, asin)} pair receives the sentinel
\texttt{recency\_days = 999}.
\textbf{Input / Output.} $9 \times 30 \to 9 \times 37$.

\begin{code}
# before
event 2022-12-27 B07MFZY2F2 Samsung SSD
# after (the 7 new columns for this event)
routine         0
novelty         1        # first time this customer buys this ASIN
recency_days  999        # sentinel: never before
cat_affinity    0.000    # 0 prior events in COMPUTER_DRIVE_OR_STORAGE
cat_count_7d    0        # nothing same-cat in last 7 days
cat_count_30d   0
brand         "samsung"  # first title token, stopwords filtered
\end{code}

\textbf{Invariants.} \texttt{validate\_state\_features}:
\texttt{recency\_days} non-negative or the 999 sentinel; all new
columns present; row count preserved.

\subsection*{3.5 Stage \textsc{temporal\_split}}

\textbf{Purpose.} Per customer, order events oldest-first and label
\texttt{split} $\in \{$\texttt{train}, \texttt{val}, \texttt{test}$\}$
with the counts derived in Sec.~2.2.
\textbf{Input / Output.} $9 \times 37 \to 9 \times 38$ (added
\texttt{split}).
\textbf{Invariants.} \texttt{validate\_split}: split values
$\subseteq \{$train, val, test$\}$; every customer has $\geq 1$ train
row.

For our one customer:
\begin{center}\small
\begin{tabular}{c l l}\toprule
$t$ & split & asin \\ \midrule
0 & train & B01JTDG084 \\
1 & train & B012JTJJJW \\
2 & train & B08RDTNVYG \\
3 & train & B07WP34LMM \\
4 & train & B089NH3Q29 \\
5 & train & B07KDDKDNN \\
6 & train & B07MFZY2F2 \\
7 & val   & B09TMHDVLD \\
8 & test  & B0BDSNHLDY \\
\bottomrule
\end{tabular}\end{center}

\subsection*{3.6 Stage \textsc{attach\_train\_popularity}}

\textbf{Purpose.} For every row, attach \texttt{popularity\_train} =
count of that ASIN in the train split. Val/test rows get the same
broadcast value; this is the train-only popularity that the LLM prompt
later buckets into the
\texttt{top 5\% / top 25\% / top 50\% / bottom 50\%} ladder.
\textbf{Input / Output.} $9 \times 38 \to 9 \times 39$.

With only one customer and nine unique ASINs, every
\texttt{popularity\_train} is either $0$ (val/test ASINs) or $1$
(train-split ASINs). The quantile ladder collapses, which is one of
the \emph{things that did not fire at N=1} we discuss in Sec.~13.
\textbf{Invariants.} WARNING-only on zero popularity (unseen outside
the train window is legal).

\subsection*{3.7 Stage \textsc{subsample} (Appendix C)}

\textbf{Purpose.} Leverage-score + KMeans customer subsample. With one
customer the mechanism is trivially no-op.
\textbf{Input / Output.} Selected IDs
\verb|["R_1lztobDYoYyZWKF"]|, per-customer weights
$\omega = [1.0]$; broadcast to per-event gives
$\omega_t = 1.0$ for all $t$.

\begin{code}
# subsample_diagnostics.json  (artifact on disk)
{
 "n_customers": 1,
 "customer_ids": ["R_1lztobDYoYyZWKF"],
 "omega_stats": {"min":1.0, "median":1.0, "max":1.0,
                 "mean":1.0, "std":0.0}
}
\end{code}

\subsection*{3.8 Stage \textsc{translate\_z\_d}}

\textbf{Purpose.} Apply the \texttt{z\_d\_mapping} block from the YAML
to produce a canonical tidy persons frame, then fit
\texttt{PersonFeatureStats} on the train split (fit-on-train). The
Amazon schema uses nine kind-handlers to produce the ten canonical
columns.
\textbf{Input / Output.} $1 \times 23$ raw persons plus train events
($7 \times 39$) $\to$ $1 \times 10$ canonical, then transformed to
$z_d \in \mathbb{R}^{p}$.

The spec claims $p = 26$. Amazon ships $p = 23$ because the
\texttt{city\_size} \texttt{external\_lookup} CSV is empty and
collapses to the single fallback label \texttt{"medium"}. The driver
reports this drop explicitly via \texttt{effective\_p},
\texttt{canonical\_p}, and \texttt{p\_reduction\_reason} in
\texttt{metrics.json}: \emph{``city\_size vocabulary reduced to 1
label(s) (canonical 4)''}.

For our customer:
\begin{code}
persons_canonical (1 row, 10 columns):
  customer_id         = "R_1lztobDYoYyZWKF"
  age_bucket          = "25-34"
  income_bucket       = "<25k"
  household_size      = 1
  has_kids            = 0          # constant sentinel (Amazon)
  city_size           = "medium"   # fallback (Amazon lookup empty)
  education           = 2          # "High school diploma or GED"
  health_rating       = 5          # 5 - (diabetes=0) - (wheelchair=0)
  risk_tolerance      = 0          # (cig=0)+(mj=0)+(alc=0)
  purchase_frequency  = 7          # count of train events
  novelty_rate        = 1.0        # all 7 train events are novel=1
\end{code}

After fitting \texttt{PersonFeatureStats} on a single-row train set,
the standardized scalars become zero (mean = value, std = 0 folded to
1.0 by policy). \texttt{log1p(7) = 2.0794}, standardized to $0$. The
resulting $z_d$ vector is:

\begin{code}
z_d = [
  0, 1, 0, 0, 0, 0,      # age_bucket one-hot (25-34)    -> 6 dims
  1, 0, 0, 0, 0,         # income_bucket one-hot (<25k)  -> 5 dims
  1, 0, 0, 0, 0,         # household_size one-hot (=1)   -> 5 dims
  0,                     # has_kids                      -> 1
  1,                     # city_size one-hot (medium)    -> 1 (collapsed)
  0.0,                   # education (standardized)      -> 1
  0.0,                   # health_rating (standardized)  -> 1
  0.0,                   # risk_tolerance (standardized) -> 1
  0.0,                   # log1p(purchase_freq) stand.   -> 1
  1.0                    # novelty_rate (pass-through)   -> 1
]
# Total: 6 + 5 + 5 + 1 + 1 + 1 + 1 + 1 + 1 + 1 = 23
\end{code}

\textbf{Invariants.} \texttt{fit\_person\_features} asserts every
\texttt{persons\_canonical} \texttt{customer\_id} appears in the train
split (fit-on-train check).

\subsection*{3.9 Stage \textsc{build\_context\_string}}

\textbf{Purpose.} Render $z_d$ as a 5--12 line human paragraph $c_d$
that the LLM sees. \texttt{suppress\_fields\_for\_c\_d()} is
\verb|("has_kids", "city_size")| on Amazon because those are constant
or lookup-collapsed. Three enrichment extras from the Wave-11
\texttt{c\_d\_extra\_fields} block are applied if present.

\textbf{Output for this customer} (verbatim, the exact string the LLM
sees):

\begin{code}
Person profile
- Age: late 20s/early 30s; household of 1, as a male.
- Income: about $15k/year.
- High-school graduate.
- Self-reports excellent health; risk-neutral.
- Buys on Amazon rarely; nearly all of orders are first-time products.
- Shops on Amazon a few times a month.
\end{code}

Every phrase comes from \texttt{DEFAULT\_PHRASINGS} in
\texttt{src/data/context\_string.py}: ``25-34'' $\to$ ``late 20s/early
30s''; ``$<$25k'' $\to$ ``about \$15k/year''; education=2 $\to$
``High-school graduate''; health\_rating=5 $\to$ ``excellent health'';
risk\_tolerance=0 $\to$ ``risk-neutral''. The gender clause
(``as a male'') comes from \texttt{Q-demos-gender = Male} via the
Wave-11 \texttt{extra\_fields} hook; the Amazon-frequency clause
(``a few times a month'') comes from
\texttt{Q-amazon-use-how-oft = Less than 5 times per month}. The
\texttt{Q-life-changes} column is NaN, so the ``Recent life event''
line is skipped.

\subsection*{3.10 Stage \textsc{build\_choice\_sets}}

\textbf{Purpose.} For every event row, produce a choice set of
$J = 10$ alternatives: the chosen ASIN plus $J - 1 = 9$ negatives
drawn from the pool of ASINs \emph{first seen strictly before} the
event's \texttt{order\_date} (temporal availability filter). Negatives
are popularity-weighted (half) and same-category (half), with the
Wave-11 dedup guard removing collisions and padding cyclically when
the pool is exhausted. The RNG seed for event $t$ resample $k$ is
\texttt{42 + 1\_000\_003 * k + t}; here $k = 0$ always (we use
\texttt{n\_resamples = 1}).
\textbf{Input / Output.} $9$ events plus adapter $\to$ $9$ records,
each a dict of
\verb|{customer_id, order_date, category, chosen_asin, choice_asins|
\verb|(list len 10), chosen_idx, chosen_features, metadata, z_d, c_d,|
\verb|alt_texts}|.

\textbf{Dedup-fallback fires on the early events.} For $t = 0$ the
available pool is empty (no prior ASINs); for $t = 1$ only
\texttt{B01JTDG084} is in the pool; and so on. The guard cycles
through the (at most one) available ASIN to fill the 9 negative slots,
logs a WARNING, and sets \texttt{metadata["dedup\_fallback"] = True}.
At scale (full Amazon, 5k customers, hundreds of thousands of ASINs)
this guard is inert.

\subsection*{3.11 Stage \textsc{alt\_text rendering}}

\textbf{Purpose.} For each of the 10 alternatives per event, render a
dict \verb|{title, category, price, popularity_rank}| that the LLM
prompt interpolates as the \texttt{ALTERNATIVE:} block. The
\texttt{popularity\_rank} string is mapped from numeric
\texttt{popularity\_train} through the
\texttt{\_popularity\_percentile\_fn} registered on the adapter:
\texttt{top 5\% / top 25\% / top 50\% / bottom 50\%}. Because the
dataset has one customer with 7 train rows, all quantiles are 1.0 and
every ASIN lands in the \texttt{"top 5\%"} bucket. At-scale this
stratifies meaningfully.

\subsection*{3.12 Stage \textsc{assemble\_batch}}

\textbf{Purpose.} Turn the list of record dicts into a tensor batch:
$z_d \in \mathbb{R}^{N \times p}$,
$E \in \mathbb{R}^{N \times J \times K \times d_e}$,
$c^* \in \mathbb{Z}^{N}$, $\omega \in \mathbb{R}^{N}$,
$\text{prices} \in \mathbb{R}^{N \times J}$, plus the diagnostic
outcomes list. For each $(i, j)$ pair we call the generator once
(\texttt{generate\_outcomes}); outcomes go through the diversity
filter; all strings then go through \texttt{encode\_batch} in a single
forward pass of the encoder; cache lookups use
\texttt{OutcomesCache(customer\_id, asin, seed, prompt\_version)} and
\texttt{EmbeddingsCache(outcome\_string, encoder\_id)}.

\textbf{Three batches.} The driver assembles
\texttt{batch\_train} ($N = 7$), \texttt{batch\_val} ($N = 1$), and
\texttt{batch\_test} ($N = 1$). Only the train $\omega$ is populated
by subsample (1.0 here); val/test pass $\omega = \text{None}$, which
\texttt{iter\_batches} interprets as uniform 1.0. Shapes:
\begin{center}\small
\begin{tabular}{l c c c}\toprule
tensor & train & val & test \\
\midrule
$z_d$             & $(7, 23)$          & $(1, 23)$          & $(1, 23)$ \\
$E$               & $(7, 10, 1, 768)$  & $(1, 10, 1, 768)$  & $(1, 10, 1, 768)$ \\
$c^*$             & $(7,)$             & $(1,)$             & $(1,)$ \\
$\omega$          & $(7,)$             & $(1,)$             & $(1,)$ \\
prices            & $(7, 10)$          & $(1, 10)$          & $(1, 10)$ \\
\bottomrule
\end{tabular}\end{center}

Note $K = 1$: this run uses the secondary ablation $K = 1$ (one
outcome per alternative). That choice has methodological consequences
detailed in Sec.~9.

\textbf{Invariants.} \texttt{assemble\_batch} fires ten shape and
sanity checks on the way out: \texttt{z\_d\_width\_uniform},
\texttt{J\_uniform}, \texttt{z\_d\_shape}, \texttt{E\_shape},
\texttt{E\_no\_nan}, \texttt{z\_d\_finite},
\texttt{c\_star\_in\_range}, \texttt{prices\_shape},
\texttt{prices\_non\_negative}, \texttt{omega\_shape}.

% ============================================================
\section*{4. Three events side by side}
% ============================================================

The pipeline treats each event uniformly, but the contents of the
choice set and the alternative rendering differ markedly depending on
how far into the timeline we are. We show three snapshots.

\subsection*{4.1 Event $t = 0$ (food scale, dedup-fallback fires)}

Available pool before 2018-02-11: empty. The dedup guard pads
cyclically with the chosen ASIN's own code (no temporally-earlier
negatives exist for this sparse fixture). Every alt-text therefore
repeats \texttt{B01JTDG084} details.
\texttt{metadata["dedup\_fallback"] = True}.

\begin{center}\footnotesize
\begin{tabular}{c l l r l}\toprule
$j$ & title (truncated) & category & price & popularity \\ \midrule
0 (*) & Greater Goods Food Scale & WEIGH\_SCALE & 9.99 & top 5\% \\
1     & Greater Goods Food Scale & WEIGH\_SCALE & 9.99 & top 5\% \\
2     & Greater Goods Food Scale & WEIGH\_SCALE & 9.99 & top 5\% \\
\multicolumn{5}{c}{...fallback pads 8 more identical rows;
 only \texttt{chosen\_idx} shuffles position...} \\
\bottomrule
\end{tabular}\end{center}

\subsection*{4.2 Event $t = 4$ (quilt, partial pool)}

Available pool before 2022-09-03: the four earlier ASINs
\{B01JTDG084, B012JTJJJW, B08RDTNVYG, B07WP34LMM\}. The guard resamples
up to five rounds and still has to pad: $\min(J, |\text{pool}| + 1)$
unique slots $= 5$ chosen from 4 pool + 1 chosen, then five more
fallback-padded.

\begin{center}\footnotesize
\begin{tabular}{c l l r l}\toprule
$j$ & title (truncated) & category & price & popularity \\ \midrule
0 (*) & Patchwork Quilt Set Full/Queen & BEDDING\_SET  & 49.99 & top 5\% \\
1     & Greater Goods Food Scale       & WEIGH\_SCALE  &  9.99 & top 5\% \\
2     & TBBSC Digital Pocket Scale     & WEIGH\_SCALE  & 10.80 & top 5\% \\
3     & \emph{(blank title)}           & Unknown       & 17.99 & top 5\% \\
4     & Walsh WSR17BK Refrigerator     & REFRIGERATOR  & 99.99 & top 5\% \\
5--9  & Patchwork Quilt Set (dedup-pad cycle) & BEDDING\_SET & 49.99 & top 5\% \\
\bottomrule
\end{tabular}\end{center}

\subsection*{4.3 Event $t = 8$ (HORI controller; test split)}

Available pool is all eight earlier ASINs. The dedup guard now has
enough distinct items to fill the choice set:

\begin{center}\footnotesize
\begin{tabular}{c l l r l}\toprule
$j$ & title (truncated) & category & price & popularity \\ \midrule
 0   & Walsh WSR17BK Refrigerator     & REFRIGERATOR                   & 99.99 & top 5\% \\
 1   & TBBSC Digital Pocket Scale     & WEIGH\_SCALE                   & 10.80 & top 5\% \\
 2   & Greater Goods Food Scale       & WEIGH\_SCALE                   &  9.99 & top 5\% \\
 3   & Pokemon Violet                 & PHYSICAL\_VIDEO\_GAME\_SOFTWARE & 54.00 & (val; rank=1) \\
 4   & Patchwork Quilt Set            & BEDDING\_SET                   & 49.99 & top 5\% \\
 5   & Greater Goods Food Scale       & WEIGH\_SCALE                   &  9.99 & top 5\% \\
 6   & (blank title)                  & Unknown                        & 17.99 & top 5\% \\
 7   & TBBSC Digital Pocket Scale     & WEIGH\_SCALE                   & 10.80 & top 5\% \\
 8   & Walsh WSR17BK Refrigerator     & REFRIGERATOR                   & 99.99 & top 5\% \\
 9 (*)& HORI Switch Split Pad Compact & VIDEO\_GAME\_CONTROLLER        & 49.99 & top 5\% \\
\bottomrule
\end{tabular}\end{center}

The exact permutation depends on
\texttt{seed + 1\_000\_003 * 0 + 8} and on the slot shuffle of the
chosen ASIN; positions above are illustrative but the identities are
exactly those that sit in the event-$t=8$ pool.

% ============================================================
\section*{5. The exact LLM prompt for one (event, alternative) pair}
% ============================================================

We show the verbatim prompt for event $t = 3$ (refrigerator, price
\$99.99), alternative $j = 0$ (same ASIN; it is the chosen one). This
is the row that produced the cached outcome \texttt{``This \$100
purchase...''} on 2026-04-21 20:03:55 UTC.

\subsection*{5.1 System prompt}

\texttt{build\_system\_prompt(K=1)} substitutes \verb|{K}| into the
frozen template from \texttt{src/outcomes/prompts.py}:

\begin{codetiny}
You generate short first-person outcome narratives for a decision-maker considering an alternative. Produce exactly 1 sentences, each 10-25 words, each describing a *different type* of consequence: financial, health/physical, convenience/time, emotional/identity, social/relational. Sentences are in the first person, present or near-future tense, and grounded in the person's specific context. Do not describe the product. Do not hedge ("might", "could" are fine; "I think" is not). Do not number the sentences; separate them with newlines.
\end{codetiny}

\subsection*{5.2 User block}

\begin{codetiny}
CONTEXT:
Person profile
- Age: late 20s/early 30s; household of 1, as a male.
- Income: about $15k/year.
- High-school graduate.
- Self-reports excellent health; risk-neutral.
- Buys on Amazon rarely; nearly all of orders are first-time products.
- Shops on Amazon a few times a month.

ALTERNATIVE:
- Name/title: Walsh WSR17BK Compact Refrigerator, 1.7 Cu.Ft Single Door Fridge, Adjustable Mechanical Thermostat with Chiller, Reversible Doors, Black
- Category: REFRIGERATOR
- Price: $99.99
- Popularity: top 5%

Generate K=1 outcome sentences.
\end{codetiny}

\subsection*{5.3 What Claude Opus 4.6 returned}

Pulled verbatim from \texttt{outcomes\_cache/outcomes.sqlite} (the row
whose JSON payload timestamp matches the refrigerator's generation
time):

\begin{codetiny}
"This $100 purchase takes a noticeable chunk from my tight monthly budget, but it saves me money on takeout by keeping fresh food at home."
\end{codetiny}

\texttt{metadata} for that cache entry:

\begin{codetiny}
{"temperature": 0.8, "top_p": 0.95, "max_tokens": 180,
 "model_id": "claude-opus-4-6", "finish_reason": "end_turn",
 "seed": 42, "prompt_version": "v1", "timestamp": 1776816235.501798}
\end{codetiny}

The narrative is exactly what the system prompt asks for: first-person,
present-tense, grounded in the person's \$15k/year context, and
consequential (``tight monthly budget... saves me money on takeout'')
rather than descriptive. Note the model does \emph{not} describe the
product beyond its dollar magnitude.

\subsection*{5.4 The other eight cached outcomes for this customer}

The \texttt{outcomes.sqlite} database holds exactly nine entries, one
per unique \texttt{(customer, asin)} pair. Listed in timestamp order:

\begin{center}\footnotesize
\begin{longtable}{@{}c l p{0.55\textwidth}@{}}
\toprule
\# & ASIN & cached outcome (verbatim) \\
\midrule\endhead
0 & B01JTDG084 & I save money on groceries by portioning ingredients precisely, stretching my tight budget further each week. \\
1 & B012JTJJJW & I save time measuring small items precisely at home instead of guessing or borrowing someone else's scale. \\
2 & B07WP34LMM & This \$100 purchase takes a noticeable chunk from my tight monthly budget, but it saves me money on takeout by keeping fresh food at home. \\
3 & B08RDTNVYG & Spending \$18 on a single item takes a noticeable bite out of my tight monthly budget, even if it's a popular choice. \\
4 & B089NH3Q29 & I feel a little brighter coming home to a cozy, fresh-looking bed that actually reflects my personal style. \\
5 & B07MFZY2F2 & My computer boots and loads games in seconds, saving me minutes of waiting every single day. \\
6 & B07KDDKDNN & My SSD runs noticeably cooler during long gaming or file-transfer sessions, giving me peace of mind about my build's longevity. \\
7 & B09TMHDVLD & Spending \$54 on a single game eats up over a third of my monthly discretionary budget, leaving me tight on essentials for weeks. \\
8 & B0BDSNHLDY & My hands feel noticeably less cramped during long handheld gaming sessions, making it easier to play for hours without discomfort. \\
\bottomrule
\end{longtable}\end{center}

The narratives read like a consistent internal monologue for the
character the prompt sketches: \emph{tight budget, single-adult
household, recurring ``is this worth it'' reasoning, pragmatism about
convenience gains on computer hardware}. Semantic clustering along
financial (B01JTDG084, B07WP34LMM, B08RDTNVYG, B09TMHDVLD) versus
convenience (B012JTJJJW, B07MFZY2F2, B0BDSNHLDY) versus
emotional-identity (B089NH3Q29) versus health (B07KDDKDNN ``peace of
mind'') is already visible by eye, which is what Sec.~12 asks of the
attribute heads.

% ============================================================
\section*{6. Encoding and cache key structure}
% ============================================================

\subsection*{6.1 The 768-dim embedding (first three coordinates)}

The encoder is \texttt{sentence-transformers/all-mpnet-base-v2},
pooling \texttt{mean}, max length 64 tokens. The nine outcome strings
are run in a single forward pass to produce nine vectors of shape
$(768,)$, L2-normalized to unit norm. For the food-scale outcome the
cached embedding starts:

\begin{code}
np.float32[-0.0506, 0.0414, -0.0229, ...  (765 more dims)]
# ||e||_2 = 1.000000  (unit-normalized per config)
# encoder_id =
#   "sentence-transformers/all-mpnet-base-v2|pooling=mean|max_length=64"
\end{code}

These are the first three values of the first row in
\texttt{embeddings\_cache/embeddings.sqlite} when iterated in
\texttt{created\_at} order.

\subsection*{6.2 Outcome cache key: the bytes}

\texttt{OutcomesCache.outcomes\_key(customer\_id, asin, seed,
prompt\_version)} composes the preimage as the NUL-separated UTF-8
encoding of the four fields, then SHA-256's it. For
\texttt{B07WP34LMM}:

\begin{code}
preimage bytes (36 bytes total):
  52 5f 31 6c 7a 74 6f 62 44 59 6f 59 79 5a 57 4b 46   # "R_1lztobDYoYyZWKF"
  00                                                    # separator
  42 30 37 57 50 33 34 4c 4d 4d                         # "B07WP34LMM"
  00
  34 32                                                 # "42"
  00
  76 31                                                 # "v1"

sha256(preimage) = 64 hex characters, 32 bytes.
\end{code}

The layout is
\texttt{customer\_id \textbackslash x00 asin \textbackslash x00 42 \textbackslash x00 v1}
for the default run, and
\texttt{... \textbackslash x00 v1-K1} for the K-swept version if the
caller bumps \texttt{prompt\_version} to isolate the cache across K
(the cache-swap convention from Sec.~3.4 of the spec). For this
dry-run, \texttt{prompt\_version = "v1"} and the cache currently holds
nine entries.

\subsection*{6.3 Embedding cache key}

\texttt{EmbeddingsCache.embedding\_key(outcome\_string, encoder\_id)}
uses the same SHA-256 scheme:

\begin{code}
preimage = outcome_string.encode("utf-8")
         + b"\x00"
         + "sentence-transformers/all-mpnet-base-v2|pooling=mean|max_length=64".encode()

sha256(preimage) = 64 hex characters, 32 bytes.
\end{code}

The two caches are orthogonal: changing the prompt template
invalidates \texttt{outcomes.sqlite} entries but not
\texttt{embeddings.sqlite} entries (string-level embeddings still
apply to any string regenerated identically). Changing the encoder
\texttt{model\_id}, \texttt{pooling}, or \texttt{max\_length}
invalidates \texttt{embeddings.sqlite} without touching
\texttt{outcomes.sqlite}.

% ============================================================
\section*{7. Model intermediates with actual shapes and values}
% ============================================================

\texttt{POLEU(M=5, K=1, J=10, d\_e=768, p=23)} has 544{,}491 trainable
parameters (the corrected Amazon-$p = 23$ total from NOTES.md). The
forward pass on the train batch ($N = 7$) produces five intermediates;
we spell them out.

\subsection*{7.1 $A$: attribute scores, shape $(7, 10, 1, 5)$}

Each of the $M = 5$ heads takes the 768-dim outcome embedding and
emits a scalar. For the first val event in
\texttt{per\_decision.json} (chosen alternative $j = 3$, the single
outcome $k = 0$), the five heads return:

\begin{code}
A[0, 3, 0, :]                  # shape (5,)
# [-0.06982, -0.03522,  0.02180,  0.01320, -0.09319]
# heads:  financial, health, convenience, emotional, social
\end{code}

This is the attribute-score row for the \emph{chosen} alternative
(Pokemon Violet, \$54) in the val choice set. The signs are not yet
informative four steps in; the magnitudes ($\pm 0.1$) are typical for
a freshly-trained MPNet-backed head. At deeper training the magnitudes
widen and the signs consolidate.

\subsection*{7.2 $w$: attribute weights, shape $(7, 5)$}

The weight-net maps $z_d \to w \in \Delta^{M-1}$ via
$\text{softmax}(\text{Linear}(23 \to 32) \circ \text{ReLU} \circ
\text{Linear}(32 \to 5))$. For this customer (from the
\texttt{per\_decision.json} ``weights'' block):

\begin{code}
w[0] = [0.1008, 0.2308, 0.1180, 0.2398, 0.3105]
#        fin     health  conv   emot    social       sum = 1.00000
\end{code}

The \texttt{social} head receives the largest share (0.310), followed
by \texttt{emotional} (0.240) and \texttt{health} (0.231). This is an
\emph{early-training} weight distribution: the softmax is close to
uniform ($1/5 = 0.20$ for each) and no single head dominates. At
convergence on a 100-customer run the distribution typically sharpens
toward one or two of the five heads per customer.

\subsection*{7.3 $U$: outcome utility, shape $(7, 10, 1)$}

$U_{t, j, k} = \sum_m w_{t, m}\, A_{t, j, k, m}$. For the val event
across all ten alternatives:

\begin{code}
U[0, :, 0] =
[-0.13988, -0.03573, -0.19354, -0.03837, -0.00076,
 -0.19354, -0.08637, -0.03573, -0.13988, -0.08637]
\end{code}

Alternative $j = 4$ has the highest $U$ ($-0.00076$, essentially
zero). With $K = 1$ only, these values equal $V$ (salience is
degenerate; see next).

\subsection*{7.4 $S$: salience, shape $(7, 10, 1)$}

Salience is a softmax over $K$. With $K = 1$, softmax of a length-1
vector is identically $1.0$: every entry of $S$ is exactly $1.0$ with
gradient zero. This is the \textbf{$K = 1$ degeneracy} the methodology
section of the paper must flag.

\begin{code}
S[0, :, 0] = [1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0, 1.0]
\end{code}

\subsection*{7.5 $V$: alternative value, shape $(7, 10)$}

$V_{t, j} = \sum_k S_{t, j, k} U_{t, j, k} = U_{t, j, 0}$ when
$K = 1$:

\begin{code}
V[0] =
[-0.13988, -0.03573, -0.19354, -0.03837, -0.00076,
 -0.19354, -0.08637, -0.03573, -0.13988, -0.08637]
\end{code}

\subsection*{7.6 Logits and probabilities}

$\text{logits}[0] = V[0] / \tau$ with $\tau = 1.0$, then
$P[0] = \text{softmax}(V[0])$:

\begin{code}
P[0] =
[0.09541, 0.10588, 0.09042, 0.10560, 0.10965,
 0.09042, 0.10065, 0.10588, 0.09541, 0.10065]
\end{code}

The probabilities sit in a tight band around $0.10 = 1/J$. At
untrained init the softmax over ten alternatives is near-uniform;
after four gradient steps there is only a $\pm 0.01$ spread. The
chosen alternative \texttt{chosen\_idx = 3} has $P = 0.10560$. The
highest probability is $0.10965$ at index $4$.

% ============================================================
\section*{8. Loss computation at the data level}
% ============================================================

The per-event cross-entropy loss at the val event shown in Sec.~7
(which \texttt{per\_decision.json} records with \texttt{chosen = 3})
is:

\begin{equation*}
\begin{aligned}
\ell_0 &= -\log P(a_3 \mid \text{context}_0)
     = -\log 0.10560408979654312 \\
     &= 2.24805832 \\
\omega_0 &= 1.0 \quad (\text{subsample disabled; uniform}) \\
\omega_0 \ell_0 &= 2.24805832 \\
\mathcal{L}_{\text{data}}(\{t = 0\}) &= \frac{\omega_0 \ell_0}{\omega_0}
                                   = 2.24805832 \\
\log J &= \log 10 = 2.30258509
\end{aligned}
\end{equation*}

This is the value written to \texttt{metrics.json} as
\texttt{nll\_val}. The loss sits very close to $\log J$, the
random-guess baseline, consistent with a model trained for only four
gradient steps: $P$ is near-uniform and any one of the ten alternatives
is about equally likely under the fresh parameters.

Writing it out step by step for reader hygiene:

\begin{code}
V = [ -0.13988, -0.03573, -0.19354, -0.03837, -0.00076,
      -0.19354, -0.08637, -0.03573, -0.13988, -0.08637 ]
Z = sum(exp(V_j)) = 9.46932     # partition function
P = exp(V) / Z
# P[3] = exp(-0.03837) / 9.46932 = 0.96234 / 9.46932 = 0.10560
loss = -log(0.10560) = 2.248058
# omega_0 = 1.0, batch size = 1 here (val batch), so L_data = 2.248058
\end{code}

% ============================================================
\section*{9. Regularizer values}
% ============================================================

Per \texttt{metrics.json}:
\texttt{regularizers\_active = \{weight\_l2:true,
salience\_entropy:true, monotonicity:true, diversity:true\}}. All four
are wired; \emph{two of them are numerically inert at $K = 1$}, and
the reporter flags this honestly.

\subsection*{9.1 Weight-net $L_2$: \texttt{weight\_l2},
$\lambda = 10^{-4}$}

$\|\phi_w\|_2^2$ sums the Frobenius norms of
\texttt{WeightNet.fc1.weight} ($23 \times 32$ for Amazon) and
\texttt{WeightNet.fc2.weight} ($32 \times 5$); biases are excluded.
Order of magnitude at Xavier-uniform init:
$\mathbb{E}\|w\|_2^2 \approx 32 \cdot 23 \cdot \frac{2}{23+32}
+ 5 \cdot 32 \cdot \frac{2}{32+5} \approx 26.8 + 8.6 = 35.4$,
so $\lambda \cdot \|\phi_w\|_2^2 \approx 10^{-4} \cdot 35.4
\approx 3.5 \cdot 10^{-3}$ at init. After four training steps it
drifts down to roughly $10^{-4}$ to $10^{-3}$ depending on the sign of
the cross-entropy gradient. Compared to the data loss $\approx 2.2$,
the weight-$L_2$ contribution is numerically negligible but non-zero.

\subsection*{9.2 Salience entropy: \texttt{salience\_entropy},
$\lambda = 10^{-3}$}

The regularizer is
$-\lambda_H \cdot \mathbb{E}_{b, j}\big[-\sum_k s_k^{(j)} \log s_k^{(j)}\big]$,
i.e.\ we \emph{subtract} $\lambda_H$ times the positive Shannon entropy
so the loss falls as spread rises. At $K = 1$, $S \equiv 1$
identically, so $H(S) = -1 \cdot \log 1 = 0$ for every $(b, j)$. The
contribution to the loss is \textbf{exactly zero} (and so is the
gradient of this term).

\emph{At-scale note.} At $K = 3$ (paper default) the entropy term
varies smoothly with $S$ and actively prevents the collapse failure
mode ``one outcome always dominates.'' This run's zero is a
methodological artifact of $K = 1$, not a failure of the regularizer.

\subsection*{9.3 Diversity: \texttt{diversity}, $\lambda = 10^{-4}$}

Diversity is
$\text{mean}_{b, j}\big[\max_{k \ne k'} \cos(e_k, e_{k'})\big]$.
With $K = 1$ there is only one outcome per alternative, so the index
set $\{(k, k') : k \ne k'\}$ is empty. The regularizer returns zero
(the implementation emits $0.0$ with \texttt{torch.zeros\_like},
keeping the backward pass connected). The ``\texttt{diversity: true}''
flag in \texttt{regularizers\_active} records that the term is
\emph{wired}, not that it is nonzero.

\subsection*{9.4 Monotonicity: \texttt{monotonicity},
$\lambda = 10^{-3}$ (\textbf{fires})}

Price monotonicity is the only non-inert optional regularizer at
$K = 1$. Its implementation (Sec.~9.2 of \texttt{redesign.md}, with
the Wave-6 alt-level surrogate) proceeds as follows on every batch:
\begin{enumerate}
\item Mean-pool $u_{\text{fin}}(e_k)$ over $K$ (trivial at $K = 1$:
pool of one) to get $\bar{u}_{\text{fin}} \in \mathbb{R}^{B \times J}$.
\item Within each batch row, sort the $J$ alternatives ascending by
\texttt{prices}.
\item For every adjacent pair $(j, j+1)$ in sorted order with
$\Delta p_{j, j+1} > 0$, compute
$r = \max(0, \Delta \bar{u}_{\text{fin}} / \Delta p)^2$.
\item Average $r$ over valid pairs and over the batch; multiply by
$\lambda_{\text{mono}}$.
\end{enumerate}
With the adapter-provided \texttt{prices} tensor threaded through
\texttt{iter\_batches}, the regularizer fires on all seven train
events. At init, $u_{\text{fin}}$ is a random function of $e$, so
sign-consistency across prices is $\approx 50\%$; the quantity
averaged is order $10^{-3}$, giving $\lambda \cdot \text{avg} \approx
10^{-6}$ contribution to the loss. Over the full training horizon
(30 epochs at scale) this term pulls the financial head's price
derivative non-positive on the aggregate, which is the domain prior
the paper wants.

\subsection*{9.5 $K = 1$ zero-out nuance, explicit}

\textbf{Methodological fact for the paper}: the default $K = 3$
activates all four regularizers; setting $K = 1$ makes two of them,
salience entropy and diversity, trivially zero. The weight-$L_2$ and
monotonicity terms remain active independent of $K$. The
\texttt{regularizers\_active} dict reports the wiring bool, so a
reader inspecting \texttt{metrics.json} must also check $K$ before
concluding a regularizer is effective on any given run.

% ============================================================
\section*{10. Training trajectory}
% ============================================================

Total steps: $4$. One epoch over $7$ training events at batch size
$2$ is $\lceil 7/2\rceil = 4$ batches. The cosine scheduler with
\texttt{T\_max = 4}, \texttt{eta\_min = 1e-4} traces:

\begin{center}\small
\begin{tabular}{c r r r}\toprule
step & lr & train batch size & approx. batch loss \\
\midrule
1 & $1.0000 \cdot 10^{-3}$ & 2 & $\approx 2.36$ \\
2 & $7.6537 \cdot 10^{-4}$ & 2 & $\approx 2.34$ \\
3 & $3.4524 \cdot 10^{-4}$ & 2 & $\approx 2.33$ \\
4 & $1.0000 \cdot 10^{-4}$ & 1 & $\approx 2.32$ \\
\bottomrule
\end{tabular}\end{center}

The intermediate batch losses are interpolated between the reported
end-of-epoch \texttt{train\_loss} $= 2.3399$ and the random-init
baseline near $\log 10 = 2.303$; the only precise figure is the final
epoch mean: \texttt{train\_state.train\_loss} $= 2.33986$ in
\texttt{smoke\_summary.json}.

End-of-epoch val NLL: \texttt{val\_nll} $= 2.24806$. No early stop
(\texttt{stopped\_early = False}; patience 5, we ran 1 epoch).
\texttt{best\_val\_nll} $= 2.24806$.

% ============================================================
\section*{11. Evaluation}
% ============================================================

The driver writes two metric files: \texttt{metrics.json} (computed
on \texttt{batch\_val}, $N = 1$) and \texttt{metrics\_test.json}
(computed on \texttt{batch\_test}, $N = 1$).

\begin{center}\small
\begin{tabular}{l r r}\toprule
metric & val ($N = 1$) & test ($N = 1$) \\
\midrule
top-1         & 0.000       & 1.000       \\
top-5         & 1.000       & 1.000       \\
MRR           & 0.250       & 1.000       \\
NLL           & 2.248       & 2.184       \\
AIC           & 1{,}089{,}013 & 1{,}089{,}013 \\
BIC           & 1{,}059{,}562 & 1{,}059{,}561 \\
\bottomrule
\end{tabular}\end{center}

$n_{\text{params}} = 544{,}491$ (Amazon-$p = 23$ corrected total);
$n_{\text{train}} = 7$. AIC $= 2k + 2 n_{\text{train}} \cdot
\text{NLL}_{\text{test}} = 2 (544491) + 14 \cdot 2.184 =
1{,}089{,}012.58$; BIC swaps $2k$ for $k \ln n_{\text{train}}$ and
gives $544491 \cdot \ln 7 + 14 \cdot 2.184 \approx 1{,}059{,}561$.

\textbf{Top-1 val $= 0$, top-1 test $= 1$: why the flip?}
Both splits are a \emph{single event}. Top-1 is either $0$ (the
model's arg-max alternative is not the chosen one) or $1$ (arg-max
\emph{is} chosen). At near-init probabilities around $1/J = 0.10$ with
a spread of $\pm 0.01$, the winner is effectively stochastic across
seeds; the fact that the val single-event arg-max lost and the test
single-event arg-max won is \emph{random}. MRR corroborates: the val
MRR is $0.25$ (the chosen alternative was ranked in position 4 out of
10; reciprocal rank $1/4$); the test MRR is $1.0$ (chosen was first).
NLL, a smooth function of the full probability vector, is a saner
scalar on $N = 1$ and agrees with the uniform-ish distribution: $2.18$
and $2.25$ are within $0.1$ of each other and both within $0.12$ of
$\log J$.

The paper must therefore report $N = 1$ top-$k$ numbers \emph{only as
sanity checks}: they have zero statistical power. The 100-customer
Wave-11 run will provide real confidence intervals on top-1 / MRR.

% ============================================================
\section*{12. Interpretability reports}
% ============================================================

Four JSON files land under \texttt{/tmp/w11\_dryrun\_final/}. We
summarize each with what it says for this customer.

\subsection*{12.1 \texttt{head\_naming.json}}

Spec asks for the top-100 outcome strings per head by $u_m(e)$. The
dry-run has only 9 unique outcome strings; the top-100 slicing
collapses to 10 rows per head (duplicates from the dedup-pad cycle).
Two representative rows per head:

\begin{center}\footnotesize
\begin{longtable}{@{}l r p{0.55\textwidth}@{}}
\toprule
head & score & outcome (truncated) \\ \midrule \endhead
\texttt{m0} financial   & $-0.056$ & ``I feel a little brighter coming home to a cozy, fresh-looking bed...'' \\
\texttt{m0} financial   & $-0.268$ & ``I save money on groceries by portioning ingredients precisely...'' \\
\texttt{m1} health      & $+0.025$ & ``I feel a little brighter coming home to a cozy, fresh-looking bed...'' \\
\texttt{m1} health      & $-0.187$ & ``I save money on groceries by portioning ingredients precisely...'' \\
\texttt{m2} convenience & $+0.022$ & ``Spending \$54 on a single game eats up over a third of my monthly...'' \\
\texttt{m2} convenience & $-0.158$ & ``This \$100 purchase takes a noticeable chunk from my tight monthly...'' \\
\texttt{m3} emotional   & $+0.013$ & ``Spending \$54 on a single game eats up over a third of my monthly...'' \\
\texttt{m3} emotional   & $-0.130$ & ``I save money on groceries by portioning ingredients precisely...'' \\
\texttt{m4} social      & $+0.030$ & ``I feel a little brighter coming home to a cozy, fresh-looking bed...'' \\
\texttt{m4} social      & $-0.236$ & ``I save money on groceries by portioning ingredients precisely...'' \\
\bottomrule
\end{longtable}\end{center}

At four gradient steps and nine distinct outcomes, head-naming is
\emph{uninformative}: every head's top outcome is one of three strings,
and the $m_0$ ``financial'' head's top outcome (``I feel a little
brighter...'') reads decidedly \emph{emotional}, not financial. This
is the \textbf{fundamental reason} head-naming must be run on the
100-customer converged model, not on a 4-step dry-run.

\subsection*{12.2 \texttt{per\_decision.json}}

Event-level three-panel report. We reproduced key slices of this
artifact in Sec.~7 and Sec.~8. The full file contains, for a single
chosen event:
\begin{itemize}
\item \texttt{outcomes}: list of $J$ length-$K$ string lists
($10 \times 1$ strings).
\item \texttt{attribute\_scores}: $(J, K, M) = (10, 1, 5)$ tensor.
\item \texttt{weights}: length-$M$ vector (per-person, $K$-averaged).
\item \texttt{salience}: $(J, K)$ matrix (all 1.0 at $K = 1$).
\item \texttt{values}: length-$J$ alternative values.
\item \texttt{probabilities}: length-$J$ softmax.
\item \texttt{chosen}: the ground-truth index.
\end{itemize}

For the val event shown in Sec.~7, \texttt{chosen = 3} (Pokemon
Violet, price \$54), and its probability under the 4-step model is
$0.10560$. The model's arg-max was index $4$ at $0.10965$
(``I feel a little brighter coming home...'', the quilt outcome from
the dedup-fill), a loss of $0.004$ in probability mass that is
statistically indistinguishable from a coin flip.

\subsection*{12.3 \texttt{dominant\_attribute.json}}

The dominant attribute is
$m^* = \arg\max_m w_m \cdot |\bar{u}_m(e_{c^*})|$, where $\bar{u}$ is
the $K$-mean of the attribute scores on the chosen alternative. On
one val row the breakdown therefore has a single bucket:

\begin{code}
{
 "by_dominant_attribute":
     {"m4": {"top1":0.0, "top5":1.0, "mrr":0.25, "nll":2.248}},
 "n_by_attribute": {"m4": 1}
}
\end{code}

$m^* = m_4$ (social) for this customer's val event: the social head
receives the largest weight ($0.310$) and its chosen-alternative score
has the largest absolute magnitude of the five. At 100-customer scale
the five buckets will populate with tens of events each and the
stratified top-1 breakdown becomes meaningful.

\subsection*{12.4 \texttt{counterfactual.json}}

Perturb $z_d[0, 0]$ by $+1.0$ (the default
\texttt{perturb-first-dim-+1} handler in
\texttt{run\_all\_reports}). Index $0$ of $z_d$ is the first slot of
the \texttt{age\_bucket} one-hot: the value for ``18-24''. At baseline
this slot is $0$ for our 25--34 year old; after perturbation it
becomes $1$, which the weight-net reads as ``now this person is both
18-24 and 25-34''. The resulting deltas (from the artifact):

\begin{center}\small
\begin{tabular}{l r}\toprule
quantity & value \\ \midrule
$\Delta w_0$ financial   & $+0.00595$ \\
$\Delta w_1$ health      & $+0.01169$ \\
$\Delta w_2$ convenience & $+0.00772$ \\
$\Delta w_3$ emotional   & $-0.01056$ \\
$\Delta w_4$ social      & $-0.01480$ \\
$\Delta S$ (each slot)   & $0.000$ ($K = 1$ degeneracy) \\
$P_{\text{chosen}}$ baseline   & $0.10560$ \\
$P_{\text{chosen}}$ counterf.  & $0.10566$ \\
$\Delta P_{\text{chosen}}$     & $+5.52 \cdot 10^{-5}$ \\
\bottomrule
\end{tabular}\end{center}

Interpretation: flipping the age-bucket signal toward ``18-24'' shifts
weight \emph{toward} financial / health / convenience and \emph{away
from} emotional / social. This is \emph{consistent} with a reasonable
prior (``younger, lower-income buyers weigh the practical consequences
more than the emotional ones''). The change in chosen probability is
tiny, $5.5 \cdot 10^{-5}$, because a 4-step model is only weakly
person-conditioned. The qualitative direction matters more than the
magnitude.

% ============================================================
\section*{13. What did not fire at $N = 1$}
% ============================================================

An honest catalog of the things this dry-run \emph{cannot} tell us,
each of which will fire at Wave-11's 100-customer scale:

\begin{enumerate}
\item \textbf{val\_nll variance.} One event $\Rightarrow$ NLL is one
draw from a wide distribution. A standard error is undefined.

\item \textbf{test\_nll variance.} Same as val; the top-1 = 1 we
report is a coin flip among ten alternatives.

\item \textbf{Stratified breakdowns.} With one val event in one
category (\texttt{PHYSICAL\_VIDEO\_GAME\_SOFTWARE}), the
category-stratified breakdown has a single bucket. So does the
repeat/novel split (the event is novel by construction: first time we
see Pokemon Violet), the time-of-day split (all 2022-12-27
transactions collapse to ``evening''), and the activity-tertile split
(only one customer exists to tertile on).

\item \textbf{Dominant-attribute stratum.} One event $\to$ one bucket
$\{m_4\}$. At scale this becomes five meaningful buckets each with
tens of events.

\item \textbf{Head naming.} Top-100 collapses to top-9 unique strings.
Sec.~12.1 already showed this is not interpretable at the dry-run
scale.

\item \textbf{Salience entropy regularizer.} Inert because $K = 1$.
Even at $K = 3$, a 4-step run gives a near-uniform $S$ and the entropy
regularizer has little work to do; the 30-epoch full run is where this
term starts pulling away from the degenerate one-hot salience
attractor.

\item \textbf{Diversity regularizer.} Inert because $K = 1$ has no
pairs to compare. Activates for $K \geq 2$.

\item \textbf{Dedup fallback.} Fires for at least the first few train
events (empty / singleton pool). At Wave-11 scale (100 customers,
thousands of unique ASINs) the fallback is exercised zero times and
the \texttt{metadata["dedup\_fallback"]} bool flips to \texttt{False}
for every record.

\item \textbf{Popularity bucketing.} Collapses to \texttt{top 5\%}
everywhere because all seven train ASINs have count $= 1$ and the
quantile ladder is degenerate. Needs at least a few dozen ASINs to
spread across the four bands.

\item \textbf{Subsample leverage.} One customer has leverage $h = 1$
trivially; the KMeans coverage and the $\omega_i \propto h_i$
proportional fill only activate once we have $\geq 2$ customers
(technically $\geq n/5$, but the interesting behavior starts at
hundreds).

\item \textbf{Early stopping.} We ran one epoch and the patience is 5;
the early-stop criterion is never consulted.

\item \textbf{Standardization statistics.} Fit on a one-row train
frame produces $\mu_{\text{education}} = 2$,
$\sigma_{\text{education}} = 0$ (folded to $1.0$), so every
standardized scalar in $z_d$ is identically $0$ for this customer. At
scale, education / health\_rating / risk\_tolerance /
purchase\_frequency contribute real person-to-person signal; here they
contribute nothing, which also means the weight-net has no interesting
variation on this customer axis.
\end{enumerate}

The dry-run's purpose is not to measure these; it is to prove the code
paths. Every cache entry, every JSON artifact, and every shape
invariant fired as designed. The numbers will only tell a real story
once we re-run the pipeline with Wave-11's hundred-customer subsample
against the warmed caches this run deposited.

% ============================================================
\section*{14. Closing summary}
% ============================================================

We walked Amazon customer \texttt{R\_1lztobDYoYyZWKF} end to end:
\begin{itemize}
\item raw CSVs (Sec.~2) through clean, join, state-features, split,
popularity, subsample, $z_d$-translate, choice-set, alt-render, and
batch-assemble (Sec.~3);
\item three representative choice sets at $t = 0, 4, 8$ showing how
dedup-fallback degrades gracefully from empty to full pools (Sec.~4);
\item a verbatim LLM prompt with the real Claude Opus 4.6 completion
for the refrigerator event (Sec.~5) and the eight other cached
narratives (Sec.~5.4);
\item the encoder output's first three dimensions and the exact
NUL-separated byte layout of both cache keys (Sec.~6);
\item the five model intermediates $A, w, U, S, V$ with real shapes
and real values (Sec.~7);
\item a step-by-step loss computation at the val event (Sec.~8);
\item all four regularizers with explicit \emph{active vs.\ inert at
$K = 1$} reasoning (Sec.~9);
\item the four-step training trajectory and six evaluation metrics
(Sec.~10--11);
\item all four interpretability reports with the real cached numbers
(Sec.~12);
\item an honest inventory of everything that did not fire because
$N = 1$ (Sec.~13).
\end{itemize}

The artifacts live at \texttt{/tmp/w11\_dryrun\_final/} (JSON
reports), \texttt{outcomes\_cache/outcomes.sqlite} (nine
\texttt{claude-opus-4-6} completions), and
\texttt{embeddings\_cache/embeddings.sqlite} (nine 768-dim
\texttt{all-mpnet-base-v2} vectors). The full Wave-11 run will replay
both caches at near-instant speed and populate every breakdown that
$N = 1$ leaves empty here.

\end{document}
